---
title: "Replication: How to Make Causal Inferences with Time-Series Cross-Sectional Data under Selection on Observables"
subtitle: "Blackwell & Glynn (2018), *American Political Science Review* 112(4): 1067–1082"
author: "Replication code"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
    fig_width: 8
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = FALSE
)
# Run the full pipeline once here so all downstream chunks have access
# to the objects created in each script.
source(here::here("R", "00_setup.R"))
source(here::here("R", "01_data_prep.R"))
source(here::here("R", "02_analysis.R"))
```

## Overview

This document replicates the analysis in:

> Blackwell, Matthew, and Adam Glynn. 2018. "How to Make Causal Inferences with Time-Series Cross-Sectional Data under Selection on Observables." *American Political Science Review* 112(4): 1067–1082. <https://doi.org/10.1017/S0003055418000357>

The paper develops methods for causal inference with TSCS panel data under **selection on observables**: situations where treatment assignment depends on observed, time-varying covariates. The key method is **inverse probability weighting (IPW)** applied within a **marginal structural model (MSM)** framework.

---

## 1  Setup — Packages

```{r packages, eval=FALSE}
# Already loaded in the setup chunk above.
# To install packages independently, run:
source(here::here("R", "00_setup.R"))
```

---

## 2  Data

```{r data, eval=FALSE}
# Data loaded in the setup chunk above.
# To re-run data preparation independently:
source(here::here("R", "01_data_prep.R"))
```

### 2.1  Data summary

```{r data-summary}
dplyr::glimpse(panel)
```

### 2.2  Treatment prevalence over time

```{r treatment-prevalence, fig.cap="Share of units under treatment (democracy) in each year."}
panel %>%
  dplyr::group_by(year) %>%
  dplyr::summarise(prop_treated = mean(democracy), .groups = "drop") %>%
  ggplot2::ggplot(ggplot2::aes(x = year, y = prop_treated)) +
  ggplot2::geom_line(colour = "steelblue", linewidth = 1) +
  ggplot2::geom_point(colour = "steelblue") +
  ggplot2::scale_y_continuous(labels = scales::percent) +
  ggplot2::labs(
    x = "Year", y = "Proportion under treatment",
    title = "Treatment prevalence over time"
  ) +
  theme_bw()
```

---

## 3  Core Analysis

```{r analysis, eval=FALSE}
# Analysis run in the setup chunk above.
# To re-run independently:
source(here::here("R", "02_analysis.R"))
```

---

## 4  Tables

### Table 1: Summary Statistics

```{r table1}
sum_stat <- panel %>%
  dplyr::select(
    `Log GDP per capita` = lgdppc,
    `Democracy (D=1)`    = democracy,
    `Education`          = education,
    `Trade Openness`     = trade,
    `Lagged Log GDP pc`  = lgdppc_lag
  ) %>%
  tidyr::pivot_longer(everything(), names_to = "Variable", values_to = "value") %>%
  dplyr::group_by(Variable) %>%
  dplyr::summarise(
    N    = dplyr::n(),
    Mean = mean(value, na.rm = TRUE),
    SD   = sd(value,   na.rm = TRUE),
    Min  = min(value,  na.rm = TRUE),
    Max  = max(value,  na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 3)))

kableExtra::kable(
  sum_stat,
  digits  = 3,
  caption = "Table 1: Summary Statistics"
) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

### Table 2: Propensity Score Model

```{r table2}
ps_coef_df <- broom::tidy(ps_denom_model, conf.int = TRUE) %>%
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 4))) %>%
  dplyr::select(
    Term           = term,
    Estimate       = estimate,
    `Std. Error`   = std.error,
    `z value`      = statistic,
    `Pr(>|z|)`     = p.value,
    `95% CI Lower` = conf.low,
    `95% CI Upper` = conf.high
  )

kableExtra::kable(
  ps_coef_df,
  digits  = 4,
  caption = "Table 2: Propensity Score Denominator Model (Logistic Regression)"
) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

### Table 3: Main Results

```{r table3}
fmt_pval <- function(p) {
  dplyr::case_when(
    p < 0.001 ~ "< 0.001",
    p < 0.01  ~ paste0(round(p, 3), "**"),
    p < 0.05  ~ paste0(round(p, 3), "*"),
    TRUE      ~ as.character(round(p, 3))
  )
}

results_formatted <- results_table %>%
  dplyr::mutate(
    `95% CI`   = paste0("[", round(Estimate - 1.96 * Std.Error, 3),
                        ", ", round(Estimate + 1.96 * Std.Error, 3), "]"),
    p.value    = fmt_pval(p.value),
    Estimate   = round(Estimate, 4),
    Std.Error  = round(Std.Error, 4)
  ) %>%
  dplyr::select(Estimator, Estimate, `Std. Error` = Std.Error, `95% CI`, `p-value` = p.value)

kableExtra::kable(
  results_formatted,
  caption = "Table 3: Effect of Democracy on Log GDP per Capita under Alternative Estimators"
) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::row_spec(nrow(results_formatted), bold = TRUE, background = "#e8f4fd")
```

> **Note:** The IPW / MSM row (highlighted) is the preferred estimator from Blackwell & Glynn (2018).  
> Standard errors are cluster-robust, clustered by country.  
> \* p < 0.05, \*\* p < 0.01.

### Table 4: Balance Diagnostics

```{r table4}
balance_formatted <- balance_df %>%
  dplyr::rename(
    Covariate            = covariate,
    `SMD (Unweighted)`   = smd_unweighted,
    `SMD (IPW Weighted)` = smd_weighted
  ) %>%
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 4)))

kableExtra::kable(
  balance_formatted,
  digits  = 4,
  caption = "Table 4: Standardised Mean Differences Before and After IPW Weighting"
) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

---

## 5  Figures

Define a shared theme:

```{r theme}
theme_replication <- function() {
  ggplot2::theme_bw(base_size = 12) +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      strip.background = ggplot2::element_rect(fill = "grey90", colour = NA),
      legend.position  = "bottom"
    )
}
```

### Figure 1: Distribution of IPW Weights

```{r figure1, fig.cap="Distribution of stabilised inverse probability weights. Values near 1 indicate little reweighting."}
ggplot2::ggplot(panel, ggplot2::aes(x = ipw_cumulative_trim)) +
  ggplot2::geom_histogram(bins = 50, fill = "steelblue", color = "white", alpha = 0.8) +
  ggplot2::geom_vline(xintercept = 1, linetype = "dashed", colour = "red", linewidth = 0.7) +
  ggplot2::labs(
    x       = "Stabilised IPW Weight (trimmed at 99th percentile)",
    y       = "Count",
    title   = "Figure 1: Distribution of Inverse Probability Weights",
    caption = "Red dashed line at weight = 1 (no reweighting)."
  ) +
  ggplot2::scale_x_continuous(labels = scales::comma) +
  theme_replication()
```

### Figure 2: Covariate Balance (Love Plot)

```{r figure2, fig.cap="Standardised mean differences before and after IPW weighting. Values inside ±0.1 (dashed lines) indicate good balance."}
balance_long <- balance_df %>%
  tidyr::pivot_longer(
    cols      = c(smd_unweighted, smd_weighted),
    names_to  = "Type",
    values_to = "SMD"
  ) %>%
  dplyr::mutate(
    Type = dplyr::recode(Type,
      smd_unweighted = "Unweighted",
      smd_weighted   = "IPW Weighted"
    ),
    Type = factor(Type, levels = c("Unweighted", "IPW Weighted")),
    covariate = dplyr::recode(covariate,
      lgdppc_lag = "Lagged Log GDP pc",
      education  = "Education",
      trade      = "Trade Openness"
    )
  )

ggplot2::ggplot(
  balance_long,
  ggplot2::aes(x = SMD, y = covariate, colour = Type, shape = Type)
) +
  ggplot2::geom_point(size = 3) +
  ggplot2::geom_vline(xintercept =  0,   linetype = "solid",  colour = "grey50") +
  ggplot2::geom_vline(xintercept =  0.1, linetype = "dashed", colour = "grey50") +
  ggplot2::geom_vline(xintercept = -0.1, linetype = "dashed", colour = "grey50") +
  ggplot2::scale_colour_manual(values = c("Unweighted" = "tomato", "IPW Weighted" = "steelblue")) +
  ggplot2::labs(
    x       = "Standardised Mean Difference",
    y       = NULL,
    colour  = NULL,
    shape   = NULL,
    title   = "Figure 2: Covariate Balance Before and After IPW Weighting",
    caption = "Dashed lines at |SMD| = 0.1 (acceptable balance threshold)."
  ) +
  theme_replication()
```

### Figure 3: Point Estimates Across Estimators

```{r figure3, fig.cap="Democracy effect estimates with 95% cluster-robust CIs. The blue point (IPW/MSM) is the preferred estimator."}
est_plot_df <- results_table %>%
  dplyr::mutate(
    ci_lo     = Estimate - 1.96 * Std.Error,
    ci_hi     = Estimate + 1.96 * Std.Error,
    Estimator = factor(
      Estimator,
      levels = c("IPW / MSM (preferred)", "First Differences",
                 "Two-Way FE", "Pooled OLS")
    )
  )

ggplot2::ggplot(
  est_plot_df,
  ggplot2::aes(x = Estimate, y = Estimator, xmin = ci_lo, xmax = ci_hi,
               colour = Estimator == "IPW / MSM (preferred)")
) +
  ggplot2::geom_pointrange(size = 0.7) +
  ggplot2::geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  ggplot2::scale_colour_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey30"),
    guide  = "none"
  ) +
  ggplot2::labs(
    x       = "Estimated Effect of Democracy on Log GDP per Capita",
    y       = NULL,
    title   = "Figure 3: Democracy Effect Estimates Across Estimators",
    caption = "Point estimates with 95% cluster-robust CIs (clustered by country).\nBlue: preferred IPW/MSM estimator."
  ) +
  theme_replication()
```

### Figure 4: Propensity Score Overlap

```{r figure4, fig.cap="Overlap in propensity scores for treated (democracy = 1) and control (democracy = 0) units. Overlap is required for IPW identification."}
ggplot2::ggplot(
  panel,
  ggplot2::aes(x = ps_denom, fill = factor(democracy), colour = factor(democracy))
) +
  ggplot2::geom_density(alpha = 0.3) +
  ggplot2::scale_fill_manual(
    values = c("0" = "tomato",    "1" = "steelblue"),
    labels = c("0" = "Non-democracy (D=0)", "1" = "Democracy (D=1)")
  ) +
  ggplot2::scale_colour_manual(
    values = c("0" = "tomato",    "1" = "steelblue"),
    labels = c("0" = "Non-democracy (D=0)", "1" = "Democracy (D=1)")
  ) +
  ggplot2::labs(
    x       = "Estimated Propensity Score",
    y       = "Density",
    fill    = NULL,
    colour  = NULL,
    title   = "Figure 4: Propensity Score Overlap",
    caption = "Good overlap (common support) is needed for IPW to work well."
  ) +
  theme_replication()
```

---

## 6  Key Methodological Notes

### 6.1  Why Naive OLS and Fixed Effects Can Be Biased

In TSCS data with time-varying confounders, standard estimators can be biased:

- **Pooled OLS** ignores unobserved unit heterogeneity and past confounders.
- **Fixed Effects** removes unit-level variation but fails when time-varying covariates are themselves *mediators* of the treatment effect (Blackwell & Glynn call this "dynamic bias").
- **First Differences** similarly fails under dynamic confounding.

### 6.2  The IPW / MSM Solution

Blackwell & Glynn propose fitting a **marginal structural model** (Robins et al. 2000) via IPW:

1. Estimate propensity scores $\hat{\pi}_{it} = \Pr(D_{it} = 1 \mid \bar{D}_{i,t-1}, \bar{X}_{it})$.
2. Construct stabilised weights:
   $$w_{it} = \prod_{s \leq t} \frac{\Pr(D_{is} \mid \bar{D}_{i,s-1})}{\hat{\pi}_{is}}$$
3. Fit a **weighted fixed-effects regression** of $Y_{it}$ on $D_{it}$ using weights $w_{it}$.

Under the **sequential ignorability** assumption (no unmeasured time-varying confounders), this estimator is consistent for the average causal effect.

### 6.3  Identification Assumption

The key assumption is **sequential ignorability** (also called "selection on observables" in the dynamic setting):

$$D_{it} \perp \!\!\! \perp Y_{it}(\bar{d}) \mid \bar{D}_{i,t-1}, \bar{X}_{it}$$

This is stronger than the usual cross-sectional selection-on-observables assumption because it must hold *at each period* given the **entire history** of covariates and treatments.

---

## 7  Session Info

```{r session-info}
sessionInfo()
```
